---
title: "Survival Analysis - Lung Cancer"
author: "Noelia Caba Martin"
date: "2026-03-17"
output:
  pdf_document: default
  html_document: default
---

## Introduction

This analysis explores survival patterns in patients with advanced lung cancer 
using the `lung` dataset from the `survival` package in R. 

The dataset contains clinical data from 228 patients, including survival time, 
censoring status, and several clinical variables.

The main objectives are:

- Estimate overall survival using Kaplan-Meier curves
- Compare survival between groups using the log-rank test
- Identify prognostic factors using Cox regression

## Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```
```{r libraries}
library(survival)
library(survminer)
library(ggplot2)
```
## Data Exploration
```{r load-data}
data(lung)
head(lung)
```
```{r data-summary}
dim(lung)
str(lung)
summary(lung)
```

The dataset contains **228 patients** and **10 variables**. 
Key variables for survival analysis are `time` (survival time in days) 
and `status` (1 = censored, 2 = dead).

Notable missing data: `meal.cal` has 47 missing values (20.6%) 
and `wt.loss` has 14 missing values (6.1%).

## Kaplan-Meier Analysis
```{r kaplan-meier}
# Create survival object
surv_object <- Surv(time = lung$time, event = lung$status == 2)

# Fit Kaplan-Meier model
km_fit <- survfit(surv_object ~ 1, data = lung)

# Summary at key timepoints
summary(km_fit, times = c(180, 365, 540, 720))
```
```{r km-plot}
ggsurvplot(
  km_fit,
  data = lung,
  risk.table = TRUE,
  pval = FALSE,
  conf.int = TRUE,
  xlab = "Time (days)",
  ylab = "Survival probability",
  title = "Overall Survival - Lung Cancer Patients",
  ggtheme = theme_minimal()
)
```

The Kaplan-Meier curve shows a steep decline in survival probability during 
the first year, with a median survival time of approximately 305 days. 

At 6 months, 72.2% of patients were still alive (95% CI: 66.5% - 78.3%).
At 1 year, survival dropped to 40.9% (95% CI: 34.5% - 48.6%).
At 2 years, only 12.5% of patients remained alive (95% CI: 7.9% - 19.7%).

The wide confidence intervals at later timepoints reflect the small number 
of patients remaining at risk, which reduces the precision of the estimates.

## Survival by Sex

A key clinical question is whether survival differs between male and female 
patients with advanced lung cancer.

**Hypothesis:**

- **H₀ (null hypothesis):** There is no difference in survival between male 
and female patients
- **H₁ (alternative hypothesis):** There is a statistically significant 
difference in survival between male and female patients

The log-rank test will be used to compare the two survival curves, with a 
significance threshold of α = 0.05.

```{r km-by-sex}
# Fit Kaplan-Meier by sex
km_sex <- survfit(surv_object ~ sex, data = lung)

# Plot
ggsurvplot(
  km_sex,
  data = lung,
  risk.table = TRUE,
  pval = TRUE,
  conf.int = TRUE,
  xlab = "Time (days)",
  ylab = "Survival probability",
  title = "Survival by Sex - Lung Cancer Patients",
  legend.labs = c("Male", "Female"),
  palette = c("#E7B800", "#2E9FDF"),
  ggtheme = theme_minimal()
)
```

While the Kaplan-Meier analysis allows us to compare survival between groups, 
it can only evaluate one variable at a time. Cox proportional hazards regression 
allows us to assess the effect of multiple variables simultaneously, adjusting 
for potential confounders.

The key output of Cox regression is the **Hazard Ratio (HR)**:

- **HR = 1:** no effect on survival
- **HR > 1:** increased risk of death
- **HR < 1:** reduced risk of death (protective factor)

**Hypothesis:**

- **H₀:** None of the included variables (age, sex, ECOG status) are associated 
with survival
- **H₁:** At least one variable is significantly associated with survival

The following variables were selected based on clinical relevance and 
data completeness:

- **Age:** continuous variable, measured in years
- **Sex:** binary variable (1 = male, 2 = female)
- **ECOG performance status:** ordinal scale measuring patient functional 
status (0 = asymptomatic, 4 = bedridden)

```{r cox-regression}
# Fit Cox model
cox_fit <- coxph(surv_object ~ age + sex + ph.ecog, data = lung)

# Summary
summary(cox_fit)
```

The Cox regression model was fitted on 227 patients (1 observation deleted 
due to missing data). A total of 164 events (deaths) were observed.

**Results:**

The model shows good overall fit, with all three global tests highly significant 
(p = 1×10⁻⁶), indicating that the included variables jointly predict survival.

- **Age** (HR = 1.011, 95% CI: 0.993 - 1.030, p = 0.232): Age was not a 
significant predictor of survival after adjusting for sex and ECOG status. 
Each additional year of age was associated with only a 1.1% increase in 
the hazard of death.

- **Sex** (HR = 0.575, 95% CI: 0.414 - 0.799, p = 0.001): Female sex was 
associated with a 42.5% reduction in the hazard of death compared to males, 
after adjusting for age and ECOG status. This is a statistically significant 
and clinically meaningful finding.

- **ECOG performance status** (HR = 1.590, 95% CI: 1.273 - 1.986, p < 0.001): 
Each one-unit increase in ECOG score was associated with a 59% increase in 
the hazard of death. This was the strongest prognostic factor in the model.

The model concordance of 0.637 indicates moderate discriminative ability, 
which is typical for clinical survival models.

**Conclusion:** After adjusting for confounders, sex and ECOG performance 
status are independent prognostic factors for survival in advanced lung cancer 
patients, while age did not reach statistical significance in this model.

The results of the Cox regression are visualized below using a **forest plot**, 
which is the standard representation in clinical publications. 

Each variable is displayed with its Hazard Ratio (point estimate) and 
95% confidence interval (horizontal line). The vertical dashed line at 
HR = 1 represents the reference value of no effect:

- Variables with the confidence interval entirely to the **left of 1** are 
protective (reduce the risk of death)
- Variables with the confidence interval entirely to the **right of 1** 
increase the risk of death
- Variables whose confidence interval **crosses 1** are not statistically 
significant.

```{r forest-plot}
ggforest(
  cox_fit,
  data = lung,
  main = "Hazard Ratios - Cox Regression Model",
  cpositions = c(0.02, 0.22, 0.4),
  fontsize = 0.9,
  refLabel = "Reference",
  noDigits = 2
)
```

The forest plot confirms the results from the Cox regression table:

- **Age** is not a significant prognostic factor (HR = 1.01, p = 0.232), 
as its confidence interval crosses the line of no effect (HR = 1)
- **Sex** is a significant protective factor (HR = 0.58, p < 0.001), 
with the entire confidence interval to the left of HR = 1
- **ECOG performance status** is the strongest risk factor (HR = 1.59, 
p < 0.001), with the entire confidence interval to the right of HR = 1

## Verification of Cox Model Assumptions

The Cox proportional hazards model relies on a key assumption: the hazard 
ratios must remain **constant over time** (proportional hazards assumption). 
If this assumption is violated for any variable, the model results for that 
variable may not be reliable.

The proportional hazards assumption is tested using **Schoenfeld residuals**. 
For each variable, the test evaluates:

- **H₀:** The proportional hazards assumption holds (HR is constant over time)
- **H₁:** The proportional hazards assumption is violated (HR changes over time)

A p-value > 0.05 indicates that the assumption is met for that variable.

```{r cox-assumptions}
# Test proportional hazards assumption
ph_test <- cox.zph(cox_fit)

# Print results
print(ph_test)

# Plot Schoenfeld residuals
ggcoxzph(ph_test)
```

The Schoenfeld residuals test confirms that the proportional hazards 
assumption is met for all variables:

- **Age** (p = 0.665): assumption holds ✓
- **Sex** (p = 0.129): assumption holds ✓
- **ECOG status** (p = 0.152): assumption holds ✓
- **Global test** (p = 0.216): the model as a whole satisfies the 
proportional hazards assumption ✓

The graphical inspection of Schoenfeld residuals confirms these results, 
with all variables showing approximately horizontal smoothed lines over 
time, indicating stable hazard ratios throughout the follow-up period.

**Conclusion:** The proportional hazards assumption is not violated, 
validating the use of the Cox model for this dataset.

## Conclusions

This analysis demonstrates the application of standard survival analysis 
methods to real clinical data from patients with advanced lung cancer.

**Key findings:**

1. The median overall survival was approximately 305 days, with only 12.5% 
of patients surviving beyond 2 years, reflecting the poor prognosis 
associated with advanced lung cancer.

2. Female patients showed significantly better survival than male patients 
(log-rank p = 0.0013), a finding that remained significant after adjusting 
for confounders in the Cox model (HR = 0.58, p < 0.001).

3. ECOG performance status was the strongest independent prognostic factor 
(HR = 1.59, p < 0.001), confirming that functional status at diagnosis is 
a critical determinant of survival outcomes.

4. Age was not a significant predictor of survival in this model 
(HR = 1.01, p = 0.232), suggesting that functional status and sex are 
more important prognostic factors than age alone in this patient population.

**Limitations:**

- The dataset contains missing values in several variables, particularly 
`meal.cal` (20.6% missing), which were excluded from the analysis
- The study population is limited to North American patients from a single 
clinical trial, which may limit generalizability
- Additional variables such as tumor histology, stage, and treatment 
information were not available in this dataset

**Future directions:**

- Include additional clinical and molecular variables to improve model 
performance
- Validate the model in an independent dataset
- Explore machine learning approaches for survival prediction